---
title: "White Wine Quality Analysis"
author: "Partho Mandal"
date: "March 9, 2017"
output: html_document
---

#Introduction
The dataset used for this project is white wine quality dataset.This data set contains information on 1,599 different red wines from a 2009 study. The dataset consists of 11 variables and 4898 observations.The aim of my investigation is to 
see if any variables affect volatile acidity which in turn affects the quality 
of the white wine. 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```


```{r packages}
# Load all of the packages 

library(ggplot2)
library(reshape)
library(corrplot)
```

```{r Load_the_Data}
# Load the Data
wine <- read.csv('wineQualityWhites.csv')
```

#Data Summary

The dataset description is shown below. We created a new variable called bound sulfur dioxide which is nothing but total sulfur dioxide subtracted by the free sulfur dioxide.
```{r Data_Summary}
#Renamed varible X to Wine.ID
wine$X <- NULL

#Created "Bound Sulphur dioxide" varible
wine <- within(wine, bound.sulfur.dioxide <- total.sulfur.dioxide - free.sulfur.dioxide )

#Structure of Data
dim(wine)

str(wine)

#summary of Data
summary(wine)
```

Now, I will be performing Univariate, Bivariate and Multivariate analysis.

#Univariate Analysis

```{r}
#Function to generate ggplots of some features

univ_cont <- function(feat) {
    ggplot(data=wine, aes_string(x = feat)) + geom_histogram()
}

uni_va <- univ_cont("volatile.acidity")
uni_ph <- univ_cont("pH")
uni_den <- univ_cont("density")
uni_alc <- univ_cont("alcohol")
uni_sul <- univ_cont("sulphates")
```

###Volatile Acidity
```{r}
#Histogram chat of volatile.acidity
plot(uni_va)
```

```{r}
#Histogram chart of volatile.acidity - outliers removed
ggplot(aes(x = volatile.acidity), data = wine)+
  geom_histogram(binwidth = 0.01)+
  coord_trans(y = 'sqrt')+
  scale_x_continuous(limits = c(0.1,0.70), breaks = seq(0.1,0.70,0.1))
```

The distribution appears unimodal with the volatile acidity peaking around 0.28.

<br>

###Quality

Is there any effect on the quality? What does this plot looks like across the categorical variables of quality.

```{r fig.height=3}
#Bar chart of quality
ggplot(aes(x = quality), data = wine)+
  geom_bar()+
  scale_x_continuous(limits = c(0,10), breaks = seq(0,10,1))
```

The majority of white wines have a quality level 5 and 6.

<br>

###pH Level
```{r}
#Bar chart of pH level
plot(uni_ph)
```

```{r}
#Histogram chart of PH level - outliers removed
ggplot(aes(x = pH), data = wine)+
  geom_histogram(binwidth = 0.01)+
  scale_x_continuous(limits = c(2.8,3.6), breaks = seq(3,3.6,0.05))
```

```{r}
#Summary chart of pH level
summary(wine$pH)

table(wine$volatile.acidity)
```

There is a peak around 3.14. The pH level is probably affected by acidity. 
Minimum level of pH is 2.720 and maximum is 3.820.

<br>

###Density
```{r}
#Histogram of density
plot(uni_den)
```

```{r}
#Histogram of density - outliers removed
ggplot(aes(x = density), data = wine)+
  geom_histogram(binwidth = 0.0002)+
  scale_x_continuous(limits = c(0.988,1.001), breaks = seq(0.988,1.001,0.001))+
  coord_trans(y = 'sqrt')
```

```{r}
#Summary data of density
summary(wine$density)
```

Density has a very small range, from 0.9871 to 1.0390

<br>

###Alcohol percentage by volume
```{r}
#Bar chart of alcohol percentage
plot(uni_alc)
```

```{r}
#Bar chart of alcohol percentage - outliers removed
ggplot(aes(x = alcohol), data = wine)+
  geom_histogram(binwidth = 0.1)+
  scale_x_continuous(limits = c(8.5,13.6), breaks = seq(8.5,13.6))+
  coord_trans(y = 'sqrt')
```

```{r}
#Summary data of alcohol percentage
summary(wine$alcohol)
```

There is a peak around 9.4, and the distribution is skewed to the right.

<br>

###Sulphates
```{r}
#Histogram of sulphates
plot(uni_sul)
```

```{r}
#Histogram of sulphates - outliers removed
ggplot(aes(x = sulphates), data = wine)+
  geom_histogram(binwidth = 0.005)+
  scale_x_continuous(limits = c(0.15,0.9), breaks = seq(0.15,0.9,0.05))+
  coord_trans(y = 'sqrt')

table(wine$sulphates)
```

There is a peak around 0.55. Distribution is skewed to the right.

The distribution appears slightly bi-modal with the sulphate concentration 
peaking around 0.38 and again at 0.5.

<br>

###Citric Acid

```{r}
#Histogram of sulphates - outliers removed
ggplot(aes(x = citric.acid), data = wine)+
  geom_histogram(binwidth = 0.005)+
  scale_x_continuous(limits = c(0.1,0.75), breaks = seq(0.1,0.75,0.05))
```

This is a square root transformed histogram plot of citric acid. There's peak in citric acid concentration around 0.30, and a sudden spike at 0.48. Distribution 
is normal.

<br>

###Bound Sulfur Dioxide

```{r}
#Histogram of sulphates - outliers removed
 ggplot(aes(x = bound.sulfur.dioxide), data = wine)+
   geom_histogram(binwidth = 5)+
   scale_x_continuous(limits = c(5,200), breaks = seq(5,200,10))+
   coord_trans(y = 'sqrt')

```

This is a square root transformed histogram plot of bound sulfur dioxide. There 
is a peak in  bound sulfur dioxide concentration around 85. Distribution is 
skewed to the right.

<br>

###Free Sulfur Dioxide

```{r}
#Histogram of sulphates - outliers removed
ggplot(aes(x = free.sulfur.dioxide), data = wine)+
  geom_histogram(binwidth = 5)+
  scale_x_continuous(limits = c(5,100), breaks = seq(5,100, 10))+
  coord_trans(y = 'sqrt')

```

This is a square root transformed histogram plot of free sulfur dioxide. There 
is a peak in  bound sulfur dioxide concentration around 30. Distribution is 
skewed to the right.

<br>


##Q & A

### What is the structure of your dataset?
Data-frame	consists of 4898 white wines of  12 original variables (Wine id, 
fixed acidity, volatile acidity, citric acid, residual sugar, chlorides, free 
sulfur dioxide, total sulfur dioxide, density, pH, sulphates, alcohol and 
quality) + 1 derived variable(Bound Sulphur dioxide). The variable quality is ordered factor variable with the following levels.

Quality: (Worst) 0, 1, ———> , 9,10 (Best)



Salient observations:

* Most white wines have a quality of 5 or 6
* Median pH level is 3.180
* Majority of white wines have between 9 and 13 percent of alcohol


### What is/are the main feature(s) of interest in your dataset?
The main feature in the data set is volatile acidity. I wanted to find out how volatile acidity increase or decrease w.r.t the quality of the white wine. I 
suspect pH and some combination of the other variables can be used to build a predictive model to grade white wines.


### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
I would like to see if the amount residual sugar increases the quality of the 
white wine, and also if there is any connection with the amount of alcohol in the wine itself.

###Did you create any new variables from existing variables in the dataset?
A new variable was created named "bound.sulfur.dioxide". It was shown in the 
summary of the data frame and was later used in the bi-variate plots section.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?
I found that the alcohol percentage distribution was right skewed compared to 
the other variables that I investigated. Most of the white wines were below 13% 
of alcohol. In most of the cases, I removed the outliers to get a better look at 
the data.


# Bivariate Analysis
```{r Bivariate_Plots}
#Correlation matrix using pearson method
round(cor(wine, method = 'pearson'),3)

```

I noticed from the Pearson correlation above that the strongest correlations 
with volatile acidity are bound sulfur dioxide and quality. The correlation coefficients are 0.157 and -0.195, respectively. Let’s look at the visual representation of the correlations.

```{r}
#Correlation plot
cm <- round(cor(wine, method = 'pearson'),3)

corrplot(cm, method = "circle")

```

We can clearly see from the size and color of the circles that volatile acidity 
has the strongest correlation with citric acid, quality, and bound sulfur 
dioxide, as stated above. Thus, the next step will be making bi-variate plot for each of the four variables

###Volatile Acidity v/s Citric Acid 

```{r}
#Jitter Plot of citric.acid vs volatile acidity
vola <- ggplot(aes(x = citric.acid, y = volatile.acidity), data = wine)

vola + geom_jitter()
```

```{r}
#Jitter Plot of citric.acid vs volatile acidity - outliers removed
vola + geom_jitter(alpha = 1/5)+
  scale_x_continuous(limits = c(0,0.75), breaks = seq(0,0.75,0.05))+
  geom_smooth()
```

The amount of volatile acidity decreases as citric acid increases. Could the 
citric acid have an effect on the taste of the white wine?

###Volatile Acidity v/s Quality 
```{r}
#Box Plot of quality vs volatile acidity
qua <- ggplot(aes(x = factor(quality), y = volatile.acidity), data = wine)

qua + geom_boxplot()+
  geom_jitter(position=position_jitter(width=.1, height=0))
```

```{r}
#Summary of quality vs volatile acidity
by(wine$volatile.acidity, wine$quality, summary)
```

The amount of volatile acidity in level 4 of quality would confirm how volatile acidity affects the taste of the wine.


###Volatile Acidity v/s Bound Sulfur Dioxide
```{r}
#Jitter Plot of bound.sulfur.dioxide vs volatile acidity
bound <- ggplot(aes(x = bound.sulfur.dioxide, y = volatile.acidity), data = wine)

bound + geom_jitter()
```

```{r}
#Jitter Plot of bound.sulfur.dioxide vs volatile acidity - outliers removed
bound + geom_jitter(alpha = 1/3)+
  scale_y_continuous(limits = c(0.10,0.9))+
  scale_x_continuous(limits = c(25,250), breaks = seq(25,250,10))+
  geom_smooth()
```

The amount of volatile acidity increases as bound sulfur dioxide increases. 


Let's also look into alcohol against quality.

```{r}
#Jitter Plot of alcohol vs quality
alc_qua <- ggplot(aes(x = quality, y = alcohol), data = wine)
alc_qua + geom_jitter()
```

```{r}
#Jitter Plot of alcohol vs quality
alc_qua <- ggplot(aes(x = jitter(quality), y = alcohol), data = wine)
alc_qua + geom_jitter(alpha = 1/3)+
  scale_x_continuous(breaks = seq(0,10,1))+
  geom_smooth()
```

Interestingly we observe a trend : as the alcohol percentage increases so do the quality. 


```{r}
#Jitter Plot of free.sulfur.dioxide vs bound.sulfur.dioxide
fb <- ggplot(aes(x = free.sulfur.dioxide, y = bound.sulfur.dioxide), data = wine)
fb + geom_jitter(alpha = 1/3)+
  scale_x_continuous(limits = c(0,80), breaks = seq(0,80,10))+
  geom_smooth()
```

Visual of bound vs free sulfur dioxide, showing a positive correlation.


##Q & A

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

Volatile acidity correlates strongly with citric acid and bound sulfur dioxide.

The amount of volatile acidity decreases as citric acid increases, but the data 
was widely spread and only showing small clusters of data.

The overlay of jitter data on top of the box plot of volatile acidity against quality create a good visual for comparison of the different qualities.

The visual for volatile acidity against bound sulfur dioxide didn't really show 
a good explanation as the data was widely spread, but did show a increase of volatile acidity when bound sulfur dioxide had increased a lot.


### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

With the new variable that I created, it show good correlation between free 
sulfur dioxide and bound sulfur dioxide. Also, alcohol against quality showed 
that as the alcohol percentage increases so do the quality. 


### What was the strongest relationship you found?

The level of volatile acidity showed a negative correlation with quality showing that the quality of white wine increased.


# Multivariate Analysis

###Citric Acid v/s Volatile Acidity factored by Quality
```{r Multivariate_Plots}
#Jitter Plot of citric.acid vs volatile acidity factored by quality
gpfq <- geom_point(aes(color = factor(quality)))
vola + gpfq + scale_color_brewer(palette = "Reds")+
  theme_dark()
```

The volatile acidity plot elaborate on the odd trends that were seen in the box plots earlier. Most quality levels 6 and above do not exceed 0.75 of volatile acidity.

###Bound Sulfur Dioxide vs Volatile Acidity factored by Quality
```{r}
#Jitter Plot of bound.sulfur.dioxide vs volatile acidity
bound + gpfq+
  scale_y_continuous(limits = c(0.10,0.9))+
  scale_x_continuous(limits = c(25,250), breaks = seq(25,250,20))+
  scale_color_brewer(palette = "Greens")+
  theme_dark()
```

Most of the  different qualities are wide spread but there does seem to be a 
large grouping  from  45-170 grams of bound sulfur dioxide.


##Q & A

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

The citric acid plot against volatile acidity showed a good correlation as the quality of white wine increased, even though the correlation was negative.

### Were there any interesting or surprising interactions between features?

Surprisingly, we see that higher quality wines are having lower bound sulfur dioxide, which can be seen by difference in shades of green in plot.

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

No.

<!-- ------ -->

# Final Plots and Summary

### Plot One
```{r Plot_One}
ggplot(aes(x = volatile.acidity), data = wine)+
  geom_histogram(binwidth = 0.01)+
  scale_x_continuous(limits = c(0.1,0.7), breaks = seq(0.1,0.7,0.1))+
  labs(list(title = "Volatile Acidity in White Wine", x = "Volatile Acidity(g/dm3)", y = "Count of White Wines"))
```


The distribution of volatile acidity appears unimodal with a curious spike 
around 0.28.

### Plot Two
```{r Plot_Two}
alc_qua + geom_jitter(alpha = 1/3)+
  scale_x_continuous(breaks = seq(0,10,1))+
  geom_smooth()+
  labs(list(title = "Quality of Alcohol in White Wine", x = "Quality(0 to 10)", y = "Alcohol (%)"))
```


The quality level of different white wines confirmed that as the level 
increased the volatile acidity was reduced.



### Plot Three
```{r Plot_Three}
vola + gpfq +
  scale_color_brewer(palette = "Blues")+
  theme_dark()+
  labs(list(title = "Quality of Volatile Acidity vs Citric Acid in White Wine", x = "Citric Acid(g/dm3)", y = "Volatile Acidity(g/dm3)", colour = "Quality of Wine"))
```


The quality of wine increases as we move towards the lower right of the plot. 
Wine seems to have better quality when citric acid is around 0.15 and volatile acidity is 0.3.

------

# Reflection
This data set contains information on 4,898 different white wines from a 2009 
study. My goal was to find which chemical properties affected the volatile 
acidity in the white wine. I started out by exploring the distribution of 
individual variables and looked for unusual behaviors in the histograms. I then calculated and plotted the correlations between volatile acidity and the 
variables. None of the correlations were above 0.5. The two variables that had relatively strong correlations were citric acidity and bound sulfur dioxide, but 
the individual correlations were not strong enough to make definitive 
conclusions with only bi-variate analysis methods. However, plotting the multivariate plot shown as Final Plot 3 showed the increase in quality with 
certain citric acidity values. One suggestion for this data set is to include storage time and storage method since these factors can influence the quality of wine as well. Further studies might include the relationship between price and quality of wine to investigate whether expensive wines lead to better quality.
